---
title: "Social Network Analysis for American Football Games "
author: "Qinhui Xu, Zezhi Zhou, Boer Feng, Piyush Kashyap"
date: "3/5/2017"
output: pdf_document
---
###
#DATA DESCRIPTION
##The file football.gml contains the network of American football games between Division IA colleges during regular season Fall 2000, as compiled by M. Girvan and M. Newman. The nodes have values that indicate to which conferences they belong.
# 1. SETUP 
##Load the package we need to use.
```{r setup, include=FALSE}
library(igraph)
```
# 2. LOAD DATA AND SET UP ATTRIBUTE
##Load the football.gml into the working environment and add some attributes to the nodes
```{r}
wd <- getwd()
setwd(wd)
football<-read.graph("football.gml",format=c("gml"))
clo = closeness(football)
btw = betweenness(football)
ev_obj = evcent(football)
eigen = ev_obj$vector
V(football)$Degree.Centrality = degree(football)
V(football)$Closeness.Centrality = clo
V(football)$Betweenness.Centrality= btw
V(football)$Eigenvalue.Centrality =eigen
V(football)$id <- V(football)
V(football)$university <- V(football)$label
V(football)$label <- V(football)
num_nodes = vcount(football)
```
# 3. SRTUCTURE OF THE NETWORK
##We will show density, average centrality of the social network here
```{r}
density <- 613/115
print("Density:", density)
btn = mean(betweenness(football))
clo = mean(closeness(football))
deg = mean(degree(football))
g_undirected <- as.undirected(football, mode='collapse')
ev_obj <- evcent(g_undirected)
eigen <- mean(ev_obj$vector)
mydf <- do.call(rbind, Map(data.frame, Closeness=clo, 
                           Degree=deg,
                           Betweenness=btn, 
                           Eigen=eigen))
return(mydf)
```

From the result, we can see that the density of the network is about 5.33, and the average closeness of the network is really low, which is only 0.01.
Though we know the meaning of all the centrality measures in gengeral, we don't know the specific meaning of these centrality measurements in this sceanrio. After doing some research about the rules of American Football Competitions, we conclude the meaning of these measurements below:

**Degree = How many times does one university have competition with other universities?**

**Betweenness = Measurement of the winning times for one university**

**Closeness = 1/ average shortest path of one university node**

**Eigen = Power of influence of your neighbour university node**

We will explain why the average closeness is so low later.

# 4. PLOT THE CENTRALITY FOR EACH NODE
```{r}
g = football
bet <- 0
for (i in 1:num_nodes){
 #print(i)
  bet[i] <- V(g)[i]$Betweenness.Centrality
}
deg <- 0
for (i in 1:num_nodes){
  deg[i] <- V(g)[i]$Degree.Centrality
}
clo <- 0
for (i in 1:num_nodes){
  clo[i] <- V(g)[i]$Closeness.Centrality
}
eig <- 0
for (i in 1:num_nodes){
  eig[i] <- V(g)[i]$Eigenvalue.Centrality
}
matrix <- rbind(bet,deg,clo,eig)
matrix <- as.matrix(matrix)
xlabbb <-seq(1:num_nodes)
bar <- barplot(matrix,col=c("lightblue","lightpink","yellow","orange"),xlab="Node ID")
axis(1, at=bar, labels=xlabbb)

legend("topleft", c("Betweeness","Degree","Closeness","Eigenvalue"), cex=0.5, bty="n", 
       fill = c("lightblue","lightpink","yellow","orange"))
```

From the barplot, we can see that the betweeness centrality of node 85 is highest. 

Node 85 is Oklahomo University, which belongs to Big Twelve Conference

# 5. SHOW THE NETWORK
```{r}
plot(football,layout=layout.fruchterman.reingold)
```

This plots shows the whole netwrok of the football data.
Since different node belongs to different conference,
we will define the color of vertice based on the value attribute of nodes,
which contains the information of conference the node belongs to.

```{r}
colrs <- c("lavender", "lavenderblush", "lavenderblush3","lemonchiffon",
           "lightblue2","lightpink",
           "plum2","rosybrown1","thistle1","aliceblue","goldenrod1","gainsboro")
V(football)$color <- colrs[V(football)$value]
plot(football, layout = layout.fruchterman.reingold, 
 vertex.color = V(football)$color, edge.color = grey(0.5), edge.arrow.mode = "-")
legend(x=-1.5, y=-1.1, c("Atlantic Coast","Big East", "Big Ten","Big Twelve",
                     "Conference USA","Independents","Mid-American",
                     "Mountain West","Pacific Ten","Southeastern",
                     "Sun Belt","Western Athletic"), pch=21,
   col="#777777", pt.bg=colrs, pt.cex=1, cex=.8, bty="n", ncol=4)
```

From this plot we can have a basic understanding of the conference. But since there are so many nodes in the plot, which makes it a little difficult to see the distribution of the conferences, we will then remove some of the conferences and keep three of them to explain why the closeness of the network is so low.

# WHY THE AVERAGE CLOSENESS IS SO LOW?
```{r}
newfootball = football
for (i in c(0,1,2,3,5,7,8,10,11)){
    newfootball <- delete.vertices(newfootball,V(newfootball)[V(newfootball)$value == i])
}
V(newfootball)$color <- colrs[V(newfootball)$value]
plot(newfootball, layout = layout.fruchterman.reingold,edge.color = grey(0.5), edge.arrow.mode = "-")
```

From this plot, we can see that basically nodes in one conference will form an area. The nodes in the center of this area will only connect with the nodes in the outer area. This situation is consistent with the rule for American Football Competition that teams in one conference compete to each other first and then the winner will compete with the teams in other conference.

We can take the node 19 in the pink area as an example. If node 19 wants to connect with node 49, it must connect to node 62 at first, then connect to 93 and 58. The logic is same for every node in the network. Therefore, the shortest average path for every node gets higher, then the closeness will get smaller and smaller.

#WHAT IS THE DISTRIBUTION OF DIFFERENT CENTRALITY?
```{r}
g=football
closeness_g = closeness(g)
btw_g = betweenness(g)
ev_obj_g = evcent(g)
eigen_g = ev_obj_g$vector
V(g)$Degree.Centrality = degree(g)
V(g)$Closeness.Centrality = closeness_g
V(g)$Betweenness.Centrality= btw_g
V(g)$Eigenvalue.Centrality =eigen_g
v_size = get.vertex.attribute(g,"Degree.Centrality")
plot(v_size, xlab="Vertex", ylab="Degree.Centrality",
                                main="The distribution of Degree Centrality")
v_size = get.vertex.attribute(g,"Closeness.Centrality")
plot(v_size, xlab="Vertex", ylab="Closeness.Centrality",
                                main="The distribution of Closeness.Centrality")
v_size = get.vertex.attribute(g,"Betweenness.Centrality")
plot(v_size, xlab="Vertex", ylab="Betweenness.Centrality",
                                main="The distribution of Betweenness.Centrality")
v_size = get.vertex.attribute(g,"Eigenvalue.Centrality")
plot(v_size, xlab="Vertex", ylab="Eigenvalue.Centrality",
                                main="The distribution of Eigenvalue.Centrality")
```

Here we can see the distribution of these four centralities.

#WHAT DOES EGIEN CENTRALITY MEAN?
## As we mentioned before, eigen centraliy means the power of the neighbor nodes. Now, we will have a more detailed look at the eigen centrality. From the plot of distribution of Egien Centrality avove, we can find that only a few nodes have eigen centrality above 0.9. So, we will make the nodes with eigen centrality higher than 0.9 look bigger.
```{r}
bound = 0.9
V(g)$size = 5
V(g)[Eigenvalue.Centrality>=bound]$size = 15
V(g)$label = NA
V(g)[Eigenvalue.Centrality>=bound]$label = V(g)[Eigenvalue.Centrality>=bound]$id
plot(g, layout = layout.fruchterman.reingold,vertex.color = V(g)$color, 
       vertex.label = V(g)$label, vertex.label.color="black",edge.color = grey(0.5),
       edge.arrow.mode = "-")
```

The nodes in bigger size are the nodes with eigen centrality higher than 0.9. We will take node 84 as an example. 

Therefore, we need to see a subnetwork that only contains node 84 and the nodes connected to node 84.
```{r}
e = E(football)[from(84)]
sub = subgraph.edges(football, as_ids(e))
plot(sub)
```

Here, we can see that node 84 is connected to node 85, so we think maybe the influence power is related to the winning times which is betweeness centrality.

Therefore, we will explore the relationship between the eigen centrality and the average betweeness of the neighbor nodes.

## First, we will define a new attribute:avg_bet for V(football)
```{r}
for (i in 1:115){
  e = E(football)[from(i)]
  sub = subgraph.edges(football, as_ids(e))
  node_num = vcount(sub)-1
  sum_bet = 0
  list = V(sub)$id
  for (node in list){
    bet = V(football)[node]$Betweenness.Centrality
    sum_bet = bet+sum_bet
  }
  sum_bet = sum_bet - V(football)[i]$Betweenness.Centrality
  avg_bet= sum_bet/node_num
  V(football)[i]$avg_bet = avg_bet
}
V(football)$avg_bet
```

Now, we can see the average betweenness centraility for all the neighbor nodes of one node.

## Then, we will do a linear regression for the avg_bet and eigen for every node.
```{r}
avg_bet = V(football)$avg_bet
node_eigen = V(football)$Eigenvalue.Centrality
slr <- lm(node_eigen~avg_bet)
summary(slr)
```

##We can see though p-value is around 0.5, so we cannot reasonably sure that there is relationship between avg_bet and node_eigen.
##So, we will define a new attribute: avg_deg for V(football)

```{r}
for (i in 1:115){
  e = E(football)[from(i)]
  sub = subgraph.edges(football, as_ids(e))
  node_num = vcount(sub)-1
  sum_deg = 0
  list = V(sub)$id
  for (node in list){
    bet = V(football)[node]$Degree.Centrality
    sum_deg = bet+sum_deg
  }
  sum_deg = sum_deg - V(football)[i]$Degree.Centrality
  avg_deg= sum_deg/node_num
  V(football)[i]$avg_deg = avg_deg
}
V(football)$avg_deg
```

Now, we can see the average betweenness centraility for all the neighbor nodes of one node.

## Then, we will do a multiple regression for the avg_bet, avg_deg and eigen for every node.
```{r}
avg_bet = V(football)$avg_bet
avg_deg = V(football)$avg_deg
node_eigen = V(football)$Eigenvalue.Centrality
mr <- lm(node_eigen~avg_bet+avg_deg)
summary(mr)
```

## Therefore, we can be reasonably sure that eigen centrality is something related to average degree and average betweenness of all the neighbor nodes. However, we can see that the p-value for avg_bet is higher than 0.05. So, we will use eta square to see how much proportion of the variance can be uniquely explained by every independent variable.
```{r}
require(heplots)
etasq(mr,anova=TRUE,partial=FALSE)
```
## From the result above, we can conclude that there is only 0.1279 of the variance can be explained by avg_bet.
##So, average degree of the neighbor nodes is the one that really matters for the eigen centrality.

#WHICH CONFERENCE IS MORE ACTIVE?
```{r}
g = football
bet <- 0
for (i in 0:11){
  bet[i+1] <- sum(V(g)[value==i]$Betweenness.Centrality)/length(V(g)[value==i]$Betweenness.Centrality)
}
deg <- 0
for (i in 0:11){
  deg[i+1] <- sum(V(g)[value==i]$Degree.Centrality)/length(V(g)[value==i]$Degree.Centrality)
}
clo <- 0
for (i in 0:11){
  clo[i+1] <- sum(V(g)[value==i]$Closeness.Centrality)/length(V(g)[value==i]$Closeness.Centrality)
}
eig <- 0
for (i in 0:11){
  eig[i+1] <- sum(V(g)[value==i]$Eigenvalue.Centrality)/length(V(g)[value==i]$Eigenvalue.Centrality)
}
  matrix <- rbind(bet,deg,clo,eig)
  matrix <- as.matrix(matrix)
  xlabbb <-seq(0:11)
  bar = barplot(matrix,col=c("lightblue","lightpink","yellow","lightgreen"))
  axis(1, at=bar, labels=xlabbb)
  legend("topleft", c("Betweeness","Degree","Closeness","Eigenvalue"), cex=0.8, bty="n", fill = c("lightblue","lightpink","yellow","lightgreen"))
```

##Here, we can see that conference five (Independents) has the highest Betweeness. Considering the competition rule of American Football, we think this conference win many of other conferences, and then compete with other conferences. Also, the winner of this conference maybe came to the last round of the competition. Following this logic, we take a look at the betweeness of every university node in the conference five.

```{r}
V(football)[value==5]$id
V(football)[value==5]$Betweenness.Centrality
```

##We can see that node 83 has a really high betweeness (top 5 nodes with high Betweenness.Centrality). We think this node should be the winner in this conference.

#WHAT DOES BLOCK MEAN IN OUR MODEL

First, let us take a look at our cluster, community and block model.

```{r}
  matrix_row_to_col <- get.adjacency(football)
  matrix_row_to_col <-as.matrix(matrix_row_to_col)
  matrix_col_to_row <- t(matrix_row_to_col)
  matrix <- rbind(matrix_row_to_col,matrix_col_to_row  )
  cors <- cor(matrix_row_to_col)
  dissimilarity <- 1 - cors
  dist <- as.dist(dissimilarity)
  hclust <- hclust(dist)
  plot(hclust)
```

This is what cluster looks like.
```{r}
com = walktrap.community(football, steps=10)
V(football)$sg = com$membership
V(football)$color = NA
V(football)$color = rainbow(max(V(football)$sg))[V(football)$sg]
plot(football, layout = layout.fruchterman.reingold, vertex.color = 
       V(football)$color, edge.color = grey(0.5), edge.arrow.mode = "-")
```

From the community plot, we can see that there are nine blocks

```{r, include=FALSE}
library(NetCluster)
library(sna)
library(network)
library(statnet)
library(ergm)
library(NetData)
library(intergraph)
```
```{r}
    matrix_row_to_col <- get.adjacency(football)
    matrix_row_to_col <-as.matrix(matrix_row_to_col)
    matrix_col_to_row <- t(matrix_row_to_col)
    matrix <- rbind(matrix_row_to_col,matrix_col_to_row  )
    cors <- cor(matrix_row_to_col)
    dissimilarity <- 1 - cors
    dist <- as.dist(dissimilarity)
    hclust <- hclust(dist)
    num_clusters = 9
    clusters <- cutree(hclust, k = num_clusters)
    cluster_cor_mat <- clusterCorr(cors,clusters)
    gcor(cluster_cor_mat, cors)
    mean <- mean(matrix_row_to_col)
    blockmodel <- blockmodel(matrix_row_to_col, clusters)
    plot(blockmodel,main="blockmodel")
    V(football)$block = clusters
    V(football)$size = 5
    V(football)[V(football)$block == 6]$size = 15
    V(football)$label = NA
    V(football)[V(football)$block == 6]$label = V(football)[V(football)$block == 6]$id
    plot(football, layout = layout.fruchterman.reingold, vertex.size = V(football)$size,
         vertex.color = V(football)$color,vertex.label = V(football)$label,
         vertex.label.color="black",edge.color = grey(0.5),edge.arrow.mode = "-")
```

##Here we can see the block model plot of our social network.

Form the plot of block model, we can see that there is sparse interaction in block 6. 
Based on the definition of block, we can know that nodes in one block have similar position and role in the whole network. Therefore, we assume that the block 6 contains nodes which lose in the first round competition in every conference. To explore this, we visualize the nodes in block 6. We can see that the nodes in block 6 are distributed across all the conferences, which can prove our assumption to some extent.

#WHICH CNOFERENCE HAS TEAMS THAT COMPETE AMONG ONE ANOTHER?
```{r}
football <- data.matrix(football)
adj = get.adjacency(football)
adj = as.matrix(adj)
value <- V(football)$value 
value <- data.frame(value)
label <- V(football)$label
label <- data.frame(label)
df <- cbind(value,label)
foot <- set_vertex_attr(football, "value", index = V(football), df[,1])
foot <- set_vertex_attr(football, "label", index = V(football), df[,2])
foot <- asNetwork(foot)
ergm_fb <- ergm(foot ~ edges)     
summary(ergm_fb)
ergm_fb2 <- ergm(foot ~ edges + nodematch("value"))
summary(ergm_fb2)
ergm_fb3 <- ergm(foot ~ edges + nodematch("value",diff=T))
summary(ergm_fb3)
exp(-2.27144)/(1+exp(-2.27144))
```

First, we do a ergm model to explore the relationship between the probability based on edge formation. 
We can see that we get a negative edge parameter(p<50%) since the network is rather sparse.
The edge parameter (-2.27144) here is the log of the edge odds.
The corresponding probability is 0.09351607: exp(-2.27144)/(1+exp(-2.27144)).

To understand that which conference has teams that play more among one another we have to look at the individual coefficient of node match values in following summary fit chart.
	
We can see that conference eight has the significant p value and highest coefficient, i.e. 5.36, among remaining nodes. 

This indicates that team under conference eight has competed more withing among one another in comparison to teams outside its conference.

